EBASE manuscript

### 2023-04-18 manu-draft.qmd
minor text edit

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.  Median NSE values across all models and each optimization period for the different subsets of the prior combinations are shown.  For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\rho$ value for the $a$ prior distribution was low or high.  The size of the points indicate the inter-quartile range of NSE values for each subset.  As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\rho$ values. Although most most of the median value are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.  For example, large differences in the median NSE values are observed for the low and high values of $\rho$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values using a high value.  A similar, but smaller effect is seen for the $\rho$ value of the prior for the $b$ parameter.  A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.  Conversely, model results are less senstive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\rho$.  This suggests that EBASE model sensitivity is most influenced by $\rho$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08. For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, the the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differneces in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE were uninformed using the same large $\rho$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).  A one day model optimization period was used for EBASE for comparability with the Odum method.

>>>>

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.  Median NSE values across all models and each optimization period for the different subsets of the prior combinations are shown.  For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\rho$ value for the $a$ prior distribution was low or high.  The size of the points indicate the inter-quartile range of NSE values for each subset.  As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\rho$ values for each parameter. Although most of the median value are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.  For example, large differences in the median NSE values are observed for the low and high values of $\rho$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values using a high value using the 7 and 30 day optimization periods.  A similar, but smaller effect is seen for the $\rho$ value of the prior for the $b$ parameter.  A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.  Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\rho$.  This suggests that EBASE model sensitivity is most influenced by $\rho$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, with the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE were uninformed using the same large $\rho$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).  A one day model optimization period was used for EBASE for comparability with the Odum method.
